Retrofitting Effect Handlers onto OCaml

### 2020-11-02 retro-concurrency.tex
Minor edits; TK comment on effects implementation

\newcommand{\kc}[1]{{\color{red} {\it [KC says: #1]}}}
\newcommand{\am}[1]{{\color{blue} {\it [AM says: #1]}}}

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (pc)
of the handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|). The exception pointer is also updated to be the current stack
pointer (|sp|). This creates a linked-list of exception handler frames on the
stack as shown in Figure~\ref{fig:stock_stack}. Raising an exception simply
sets |sp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps to the |pc| of the
handler.

@@

In this section, we present the implementation details of effect handlers in
Multicore OCaml. The implementation follows the operational semantics, but has
a few key representational differences. Multicore OCaml currently only supports
the x86-64 backend on Linux, but our design preclude other backends. The
description in the sequel assumes we are on an x86-64 architecture.

@@

We have the variable-sized area for the OCaml frames next. Since we want the
handlers to be as small as possible, this area is initially 16 words in length.
When the stack pointer |rsp| goes less than stack threshold (a field in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks in function prologue of OCaml functions. We
observed that most function calls are to leaf functions with small frame size.
In order to take advantage of this fact, we introduce a small red zone at the
top of the stack. The compiler elides the stack overflow check for leaf
functions whose frame size is less than the size of the red zone.

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fiber only involves saving the exception and the stack
pointer of the current stack and loading the same on the target stack. Since
OCaml does not generate pointers into the stack, the two |fiber_info| fields
are the only ones that need to be updated when fibers are moved. We allocate
the space for the fiber using |malloc| and |free| it when no longer required.
We do not use a \emph{stack cache} of recently freed stacks for speeding up
fiber allocation. In our experiments, we observed that directly using
|mimalloc| outperforms having the stack cache.

@@

Since C functions do not have stack overflow checks, we have to execute the
external calls in the system stack. Calling a C function from OCaml involves
saving the exception pointer and the stack pointer in the |fiber_info| of the
current fiber, and the allocation pointer in the |Caml_state|, updating |rsp|
to the top of system stack (maintained in |Caml_state|), and calling the C
function. The actions are reversed when returning from the external call. One
subtlety is that for C functions that take arguments on the stack, the
arguments must be copied to the C stack from the OCaml stack where they are
laid out. For vast majority of external calls, we perform only a few additional
instructions stock OCaml in order to to set up the right stack.

@@

@@

For stack walking during GC, the only change necessary was to make the runtime
aware of fibers layout and how to follow the parent pointers. The stack map
generation and scanning each frame for roots remains the same. Due to the
uniformity of fiber layout, suspended fibers are scanned the same way as the
current active fiber. Previous work~\cite{RetroParallel} presents the
challenges and solution for the interaction of fibers and the concurrent
mark-and-sweep GC of Multicore OCaml.

@@

Our goal is to help DWARF identify the CFA when stacks are switched at external
calls, callbacks and effect handlers. We have implemented short, hand-written
DWARF bytecode at these three places, which has been validated with the DWARF
unwind table validator tool~\cite{Bastian}. When starting to run on a new
fiber, we insert DWARF bytecode to use the follow the |parent_fiber| pointer
and dereference the |saved_sp| to get the CFA. During callback, we save the
current system stack pointer in the DWARF block in Figure~\ref{fig:fiber} to
identify the CFA in the C stack. DWARF unwinding for external calls is
implemented similarly by saving the current OCaml stack pointer. With these
changes, we get the same backtrace for the |meander.exe| program from
\S\ref{sec:unwind}, modulo runtime system functions due to effect handlers.

>>>>

\newcommand{\kc}[1]{{\color{red} {\it [KC says: #1]}}}
\newcommand{\am}[1]{{\color{blue} {\it [AM says: #1]}}}
\newcommand{\tk}[1]{{\color{blue} {\it [TK says: #1]}}}

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (|pc|)
of the handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|). The exception pointer is also updated to be the current stack
pointer (|sp|). This creates a linked-list of exception handler frames on the
stack as shown in Figure~\ref{fig:stock_stack}. Raising an exception simply
sets |sp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps to the |pc| of the
handler.

@@

In this section, we present the implementation details of effect handlers in
Multicore OCaml. The implementation follows the operational semantics, but has
a few key representational differences. Multicore OCaml currently only supports
the x86-64 backend on Linux, but our design doesn't preclude other backends.
The description that follows assumes a x86-64 architecture.

@@

We have the variable-sized area for the OCaml frames next. Since we want the
handlers to be as small as possible, this area is initially 16 words in length.
When the stack pointer |rsp| goes less than stack threshold (a field in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions. We
observed that most function calls are to leaf functions with small frame size.
In order to take advantage of this fact, we introduce a small red zone at the
top of the stack. The compiler elides the stack overflow check for leaf
functions whose frame size is less than the size of the red zone.

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the stack
pointer of the current stack and loading the same on the target stack. Since
OCaml does not generate pointers into the stack, the two |fiber_info| fields
are the only ones that need to be updated when fibers are moved. We allocate
the space for the fiber using |malloc| and |free| it when no longer required.
We do not use a \emph{stack cache} of recently freed stacks for speeding up
fiber allocation. In our experiments, we observed that directly using
|mimalloc| outperforms having the stack cache.

@@

Since C functions do not have stack overflow checks, we have to execute the
external calls in the system stack. Calling a C function from OCaml involves
saving the exception pointer and the stack pointer in the |fiber_info| of the
current fiber, and the allocation pointer in the |Caml_state|, updating |rsp|
to the top of system stack (maintained in |Caml_state|), and calling the C
function. The actions are reversed when returning from the external call. One
subtlety is that for C functions that take arguments on the stack, the
arguments must be copied to the C stack from the OCaml stack where they are
laid out. For vast majority of external calls, we perform only a few additional
instructions over stock OCaml in order to to set up the right stack.

@@

\tk{Something wasn't quite right with this: we allocate the continuation object
at the point of the perform. Also should we describe how the first handler
passes the effect on and adds the e onto the continuation that started with f?}

@@

For stack walking during GC, the only change necessary was to make the runtime
aware of fibers layout, the closures for the handlers and how to follow the
parent pointers. The stack map generation and scanning each frame for roots
remains the same. Due to the uniformity of fiber layout, suspended fibers are
scanned the same way as the current active fiber. Previous
work~\cite{RetroParallel} presents the challenges and solution for the
interaction of fibers and the concurrent mark-and-sweep GC of Multicore OCaml.

@@

Our goal is to help DWARF identify the CFA when stacks are switched at external
calls, callbacks and effect handlers. We have implemented short, hand-written
DWARF bytecode at these three places, which we have validated with the DWARF
unwind table validator tool~\cite{Bastian}. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer
and dereference the |saved_sp| to get the CFA. During callback, we save the
current system stack pointer in the DWARF block in Figure~\ref{fig:fiber} to
identify the CFA in the C stack. DWARF unwinding for external calls is
implemented similarly by saving the current OCaml stack pointer. With these
changes, we get the same backtrace for the |meander.exe| program from
\S\ref{sec:unwind}, modulo runtime system functions due to effect handlers.
